---
title: "Packages installation"
authors: "Max Olmos & Baptiste Alglave"
date: "02/20/2024"
editor: visual
theme:
  dark: darkly
  light: flatly
highlight-style: github
format:
  html:
    documentclass: article
    papersize: letter
    toc: true
    number-sections: true
    colorlinks: true
    geometry:
      - top=30mm
      - left=20mm
      - heightrounded
    code-annotations: false
    code-tools: true
    code-block-bg: true
    code-block-border-left: "#31BAE9"
    embed-resources: true
---

------------------------------------------------------------------------

# Some info

Please make sure you have the correct version of R and packages installed. Please make sure all packages works well on your computer.

We will not take time to install packages during the workshop, we expect all participants to have VAST installed and running.

If you are facing issues while installing R or packages please contact Baptiste [baptiste.alglave\@univ-ubs.fr](baptiste.alglave@univ-ubs.fr), [**before**]{.underline} the workshop

# R and Rstudio versions

## R

R version 4.3.2 (Eye Holes). You can find it [here](https://cran.r-project.org/src/base/R-4/)

*The latest R version is R-4.3.3 (Angel Food Cakeo). But they made it available while we were preparing the workshop on February 29 2024. So please use R 4.3.2.*

## Rstudio version

Version RStudio-2023.12.1-402.exe

# Packages

## VAST

First, install the "devtools" package from CRAN

```{r}
#| eval: false

# Install and load devtools package
install.packages("devtools")
library("devtools")

```

Next, please install the VAST package from the VAST GitHub repository using a function in the "devtools" package. This may require using the INSTALL_opts option depending upon your version of R:

```{r}
#| eval: false

# Install package
install_github("james-thorson/VAST@main", INSTALL_opts="--no-staged-install")
# Load package
library(VAST)

```

If you are having problems with installation, please consider installing dependencies individually, e.g. using:

```{r}
#| eval: false
# Install TMB from CRAN
install.packages("TMB")
# Install INLA using currently recommended method
install.packages("INLA", repos=c(getOption("repos"), INLA="https://inla.r-inla-download.org/R/stable"), dep=TRUE)
# Install FishStatsUtils from CRAN
install_github("james-thorson/FishStatsUtils@main", INSTALL_opts="--no-staged-install")

```

Finally, please confirm that VAST is installed by running a model, e.g., following the simple example .

```{r}
#| eval: false

# Load package
library(VAST)

# load data set
# see `?load_example` for list of stocks with example data 
# that are installed automatically with `FishStatsUtils`. 
example = load_example( data_set="EBS_pollock" )

# Make settings (turning off bias.correct to save time for example)
settings = make_settings( n_x = 100, 
  Region = example$Region, 
  purpose = "index2", 
  bias.correct = FALSE )

# Run model
fit = fit_model( settings = settings, 
  Lat_i = example$sampling_data[,'Lat'], 
  Lon_i = example$sampling_data[,'Lon'], 
  t_i = example$sampling_data[,'Year'], 
  b_i = example$sampling_data[,'Catch_KG'], 
  a_i = example$sampling_data[,'AreaSwept_km2'] )

# Plot results
plot( fit )

```

More details about installation instructions can be found [here](https://github.com/James-Thorson-NOAA/VAST?tab=readme-ov-file#installation-instructions)

## Other packages

```{r}
#| eval: false

library(ggplot2)
library(dplyr)
library(viridis)
library(rnaturalearth)
library(rnaturalearthdata)
library(mapdata)
library(sf)
library(here)
```
